Two applications of logic programming to Coq

### 2020-11-05 papers/types2020/sec3.tex
some typos fixed in 1 2 and 3

\todo{Anything here?}
% \begin{metanote}
%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% \end{metanote}

\subsection{Proof systems with certificates}
\label{ssec:focused}
\todo{Change title? }
\todo{Use same names in Fig and 3: A for Atom, L for ListArgs etc.}

@@

Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgments for the two categories. In our case, where proof search steps are
viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
step while the judgment $\XXi\vdash l \colon D \Downarrow A$ corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

The proof system is parameterized by Coq' global environment $E$, which
here can be thought to consist of a set of inductive definitions,
which following Coq's reference manual we denote as
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  The first rule delegates
to Coq's type checking enforcing the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means, at the
term level, that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.

The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in a explicitly focused system. From a logic
programming viewpoint given an atom, it selects a clause on which to
back-chain on: we look-up the costructors from an \lsti{Inductive}
definition from the global enviroment, one that matches the atom we
atom we wish to bachchain on and then call the latter judgment. The rules for backchaining are standard.

It may be at first surprising that there are no
introduction rules for propositional connectives, nor equality for
that matter. However, one of the beauty of the Calculus of Inductive
Construction is that they are in fact defined inductively and
therefore the \emph{unfold} rule will handle those; this also
simplifies proof-terms.

@@

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
\lsti{check $\XXi$ (go A T)} and $\XXi\vdash l \colon D \Downarrow A$ corresponds to
\lsti{check $\XXi$ (bc D A L)}.

@@

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of signatures of the notions that we use:
\begin{lstlisting}
  kind term   type.   % reification of Coq terms
  kind gref    type     % reification of refernces to global terms
  type global gref -> term.
  type indt   inductive -> gref.  % reification of inductive types
  type indc   constructor -> gref. % reification of their constructors 
  type app    list term -> term.  % reification of nary application
  type prod   name -> term -> (term -> term) -> term.
\end{lstlisting}

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

\smallskip
In figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC. For example the FPC can provide a \emph{trace} that may be
more customable that the one offered by the \lsti{auto}'s hard-wired
\lsti{Debug} facility.
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\todo{Right now there is no intro to FPC}
% \begin{metanote}
%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% \end{metanote}

\subsection{Proof systems with certificates}
\label{ssec:focused}
\todo{Change title? }

@@

Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgments for the two categories. In our case, where proof search steps are
viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
step while the judgment $\XXi \Downarrow  l \colon D\vdash A$ corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

The proof system is parameterized by Coq' global environment $E$, which
here can be thought to consist of a set of inductive definitions,
which following Coq's reference manual we denote with
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
to Coq's type checking enforcing the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means, at the
term level, that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.

The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in an explicitly focused system. From a logic
programming viewpoint, given an atom, the rule selects a clause on which to
back-chain on: we look-up the costructors from an \lsti{Inductive}
definition from the global enviroment, one that matches the atom we wish to bachchain on and then call the latter judgment. The rules for backchaining are standard, corresponding to the \emph{init} and $\forall$-left rules.

It may be at first surprising that there are no
introduction rules for propositional connectives, nor equality for
that matter. However, one of the beauty of the Calculus of Inductive
Construction is that they are in fact defined inductively and
therefore the \emph{unfold} rule will handle those; this also
simplifies the syntax of proof-terms.

@@

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
\lsti{check $\XXi$ (go A T)} and $\XXi \Downarrow  l \colon D\vdash A$ corresponds to
\lsti{check $\XXi$ (bc D A L)}.

@@

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of signatures of the notions that we use:
\begin{lstlisting}
  kind term   type.                                   % reification of Coq terms
  kind gref    type                                 % reification of refernces to global terms
  type global gref -> term.                      % coercion to term
  type indt   inductive -> gref.                 % reification of inductive types
  type indc   constructor -> gref.             % reification of their constructors 
  type app    list term -> term.               % reification of nary application
  type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms: namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

In Figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC, for example iterative-deepening. Furthermore, the FPC can
provide a \emph{trace} that may be more customable that the one
offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
